Integration of national demographic-disturbance relationships into local caribou population viability projections can reduce uncertainty and inform monitoring decisions

### 2024-06-17 analysis/paper/paper.Rmd
Updated results. Draft to MH.

To derive a composition survey bias correction term we assume each group of animals in a calf:cow composition survey contains one or more collared adult females ($T$), and may also include: uncollared adult females misidentified as young bulls or unknown sex ($U$); correctly identified uncollared adult females ($V$); young bulls correctly identified as male or unknown sex ($O$); young bulls misidentified as uncollared adult females ($P$); observed calves ($J$); and unobserved calves ($K$). The apparent number of adult females in the group is $T+V+P=Tw$, where $w$ is a multiplier that defines the apparent number of adult females as a function of the number of collared animals. The ratio of young bulls to uncollared adult females in the group is
$q = (P+O)/(U+V).$ Assuming an equal probability $u$ of misidentifying young bulls as adult females and vice versa, we get $V=(U+V)(1-u)$ and $P=(O+P)u$. Given a probability $z$ of missing calves, we get $J=(J+K)(1-z)$. Given these assumptions and definitions, the sex and bias-corrected recruitment rate $X$ can be written as function of the observed calf:cow ratio $R$, the cow multiplier $w$, the ratio of young bulls to adult females $q$, and the misidentification probabilities $u$ and $z$ (see Supplement B for details). For simplicity, we combine the probabilities into a bias correction term $c$ that is integrated into the Bayesian population model and simulations (below): 
$$c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}; X=cR/2$$

@@

@@

We focus on exploring a set of scenarios in which
anthropogenic disturbance, measured as the percentage of area within 500
m of disturbance [@johnson_science_2020; @eccc_scientific_2011],
increases by 1% per year (Figure \@ref(fig:figDisturbance)). The duration of monitoring prior to the projection period varies among scenarios from 1 to 24 years. The projection period begins in 2024, and we consider how uncertainty changes as the projection period increases from 5 to 20 years. The amount of anthropogenic disturbance at the beginning of the projection period varies from low (0\%) through med-low (20\%) and med-high (40\%) to high (60\%). These simple hypothetical scenarios span a large range to show how the need for local monitoring to reduce uncertainty depends on landscape state, not to represent the reality of landscape change. Scenarios in which anthropogenic disturbance remains the same or decreases over time are included in Supplement D.

@@

To investigate the effectiveness of various monitoring strategies for reducing uncertainty about population viability, and to show how this varies with landscape condition, we begin with a model for simulating example population dynamics and observations of those dynamics. These example population trajectories are informed by prior information about the distribution of possible parameter values and outcomes from the national analysis of demographic-disturbance relationships [@johnson_science_2020] so we call it the "national" model. Our method for selecting parameter values for each example trajectory yields a continuous set of alternative models (hypotheses) about the state of a particular caribou population, wherein there are more examples of parameter values that prior analysis suggests are more likely; these alternative scenarios informed by prior beliefs are required for analysis of the expected value of alternative sampling strategies [@runge_which_2011; @dunham_strategic_2023]. We also use the simulated trajectories to confirm that our Bayesian model successfully integrates local data with priors from the national model. For reference, all variables and parameters are defined in supplemental Table A1.

@@

To simulate demographic trajectories for example populations
we used a modified version of Johnson et al's [-@johnson_science_2020]
demographic model described by Dyson et al.
[-@dyson_existing_2022]. Given a population of post-juvenile females at the beginning of year $t$, $\dot{N}_t$, the number that survive to the census, $\dot{W}_t$, is binomially
distributed with true survival probability $\dot{S}_t$:
$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. The calf:cow ratio $\dot{R}_t$ is adjusted for sex ratio, misidentification
biases (Supplement B) and (optionally) a delayed reproduction term to account for the reality that yearlings generally do not breed
[@eacker_web-based_2019; @decesare_estimating_2012] to get maximum potential recruitment of female calves per cow $\dot{X}_t$:
\begin{equation}
  \dot{X}_t=\frac{\dot{c}\dot{R}_t/2}{1+\dot{c}\dot{R}_t/2};\dot{c}=\frac{w(1+\dot{q}\dot{u}-\dot{u})}{(w+\dot{q}\dot{u}-\dot{u})(1-\dot{z})},
  (\#eq:dotX)
\end{equation}
where $w$ is a multiplier that defines the apparent number of adult
females in the composition survey as a function of the number of
collared animals, $\dot{q}$ is the ratio of young bulls to adult females, $\dot{u}$
is the probability of misidentifying young bulls as adult females and
vice versa, and $\dot{z}$ is the probability of missing a calf. We examine sensitivity to these unknown parameters by selecting values for each simulated example population from plausible ranges: $\dot{u}$ and $\dot{v}$ vary uniformly between 0 and 0.2, and $\dot{q}$ varies uniformly between 0 and 0.6.

Realized recruitment rate varies with population density
[@lacy_vortex_2017], and the number of juveniles recruiting to the
post-juvenile class at the census is a binomially distributed function
of the number of surviving post-juvenile females and the adjusted
recruitment rate:
\begin{equation}
  \dot{J}_{t} \sim \text{Binomial}(\dot{W}_t,\dot{X}_t[p_0-(p_0-p_k)(\frac{\dot{W}_t}{N_0k})^b]\frac{\dot{W}_t}{\dot{W}_t+a}).
  (\#eq:dotJ)
\end{equation}

@@

Following Dunham et al. [-@dunham_strategic_2023] we also calculate the expected value of sample information (EVSI) and the expected value of perfect information (EVPI) [@runge_which_2011;@yokota_value_2004], assigning a value of 1 to correct status assessments, and a value of 0 to incorrect assessments ($\nu_{\alpha,\psi}=1 \text{ when } \alpha=\psi, 0 \text{ otherwise}$). The expected value with perfect knowledge is $EV_{certainty}=1$. The difference between this and the expected score using knowledge of disturbance the disturbance scenario $\tau$ (Fig \@ref(fig:figDisturbance)) and priors from the national model gives EVPI, a measure of the maximum possible benefit that can be gained from local sampling:
\begin{equation}
EVPI_{\tau} = 1-max_{\alpha}\sum_\psi[P(\psi|\tau)\nu_{\alpha,\psi}],
  (\#eq:EVPI)
\end{equation}
where $P(\psi|\tau)$ is the prior probability that the true state of the population is $\psi$, and each population is assigned a status $\alpha$ informed by the prior. EVSI measures expected improvement from a particular sampling strategy as the difference between the expected value with and without additional sample information: 
\begin{equation}
EVSI_{\tau} = E_x[max_{\alpha}\sum_\psi[P(\psi|x,\tau)\nu_{\alpha,\psi}]] - (1-EVPI_{\tau}),
  (\#eq:EVSI)
\end{equation}
where $x$ is the sample information from a particular example trajectory. Simulated observations from 500 example trajectories for each disturbance and monitoring scenario $\tau$ (Figure \@ref(fig:figDisturbance)) represent the distribution of possible values of $x$. $P(\psi|x,\tau)$ is the posterior probability that the true state of the population is $\psi$, and each population is assigned a status $\alpha$ informed by the sample information.

@@

Analysis of simulated example demographic trajectories confirms that our Bayesian model can effectively combine local demographic data with priors from national demographic-disturbance relationships to reduce uncertainty in boreal caribou demographic projections (Figs \@ref(fig:examplePredictions) and \@ref(fig:figDiffs)). As expected, uncertainty about the population growth rate decreases as sampling effort in each year (i.e. number of collars) or the number of years of monitoring (Fig \@ref(fig:figDiffs)) increase. Uncertainty is higher when the forecast horizon is long (compare 5 and 20 year projections in Fig \@ref(fig:figDiffs)). Uncertainty is also higher in our high anthropogenic disturbance scenarios (Fig \@ref(fig:figDiffs)) because stochastic variation in population growth rates increase as populations become very small (as in top right panel of Fig \@ref(fig:examplePredictions)). These results confirm the Bayesian population model is working as it should.

Our analysis becomes more interesting and informative when we examine the probability of correctly assessing population status (Fig \@ref(fig:figStatusErrors)) and the expected value of information (Figs \@ref(fig:figEVSI) and \@ref(fig:figEVPI)). When anthropogenic disturbance is high (right panels of Figs \@ref(fig:figStatusErrors), \@ref(fig:figEVSI), and \@ref(fig:figEVPI)), projections of population status informed only by priors from the national demographic-disturbance relationships are accurate, and adding additional local monitoring data does not improve accuracy. Interestingly, when the projection period is long and anthropogenic disturbance is high, we see a negative EVSI (lower right panel of Fig \@ref(fig:figEVSI)), and we do not see an associated decrease in the probability of correctly assessing population status (lower right panel of Fig \@ref(fig:figStatusErrors)). This means that the long-term projections informed by local monitoring data are on average less precise but not less accurate, which can happen when there is very small prior probability that population is viable (orange band in the top right panel of Fig \@ref(fig:examplePredictions) falls below 1). At lower levels of disturbance, local monitoring can reduce uncertainty, but there are diminishing returns as the number of collars deployed or the number of years of monitoring increases (Figs \@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)). The shapes of the relationships between monitoring effort and the probability of correct status assessment depend on how anthropogenic disturbance is changing over time, but there are diminishing benefits of increased monitoring regardless of whether disturbance is increasing or decreasing (supplement D). Adding unknown bias increases uncertainty, but these increases are not large for the set of parameters we investigate (supplement E).

```{r examplePredictions, fig.cap="Prior (top row) and posterior (bottom row) predictions from the Bayesian model for an example population trajectory. In this example scenario anthropogenic disturbance increases by 2\\% per year from 0 to 100\\%. The population is monitored for 16 years with 30 collars per year, and 6 cows per collared cow in the composition surveys. The prior means and 95\\% predictive intervals (top row, orange lines and shading) are similar to the means and ranges between the 2.5\\% and 97.5\\% quantiles of 3000 simulated survival and recruitment trajectories from the national model (top row, blue lines and shading). Local population data (open triangles) reduces uncertainty (bottom row), but when disturbance is very high local data is not required to predict viability (population growth <= 0.99). When projected populations become very small, stochastic variation in projected population growth rate increases (as in the top right panel).", cache=TRUE}

@@

```{r figEVPI, fig.cap="The expected value of perfect information (EVPI) varies among anthropogenic disturbance scenarios (see increasing, decreasing and stable scenarios in Figs \\@ref(fig:figDisturbance), D.1 and D.5 respectively) and the number of years projected. When disturbance is high EVPI is low, indicating that local sampling will not alter status assessments. Increasing the forecast horizon can either increase or decrease the EVPI as the anthropogenic disturbance trend makes outcomes more or less predictable.", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/EVPIall.png"))
```

@@

Our model assumes that increasing cumulative anthropogenic disturbance causes a log-linear decrease in caribou demographic rates (equations 4, 5, 10 and 13), reflecting a scientific consensus informed by a number of lines of evidence that cumulative anthropogenic disturbance has a negative impact on boreal caribou [@eccc_scientific_2011; @johnson_science_2020; @gouvernement_du_quebec_revue_2021]. Local data can be integrated to estimate local demographic-disturbance parameters that differ from the national means, and in the absence of local data, appropriate priors lead to a range of outcomes consistent with the observed range of variation across the country [Figures 3 and 5 in @johnson_science_2020]. Others [@moeller_integrating_2021; @mcnay_demographic_2022; @eacker_web-based_2019] have also used Bayesian models to estimate caribou demographic rates from local monitoring data, but our addition of prior information from national demographic-disturbance relationships is novel, and important for reducing the risk that lack of local data will be used as a reason to delay conservation action for highly disturbed and poorly monitored populations.  Integrating demographic-disturbance relationships into a viability projection model clarifies that no additional information is needed to project outcomes where disturbance is high and actions to support recovery are not taken.

@@

Variation in anthropogenic disturbance across the Canadian boreal region is partially correlated with variation in climate and primary productivity because forestry occurs in higher productivity areas, and petroleum extraction activities are concentrated in Western Canada; concerns have been raised about confounding of demographic-disturbance relationships [@wilson_causal_2021;@neufeld_primary_2021; @neilson_direct_2022]. One option for addressing these concerns would be to develop regional demographic-disturbance relationships for populations that are ecologically similar to one another (e.g. caribou inhabiting forests that are sufficiently productive for industrial forestry in Eastern Canada) [@eccc_scientific_2024], and use the priors from regional demographic-disturbance relationships to inform demographic projections within the region. However, while acknowledging and allowing for regional variation, we also need to guard against risks that uncertainty will be inflated and used as an excuse to delay action, a tactic that has been used in this and other decision contexts [@bissonnette_controverses_2020; @boan_climate_2018]. In order to be informative and credible, regional demographic-disturbance analyses must include a sufficient number of observations spanning a range from high to low anthropogenic disturbance. Priors that lead to projections of outcomes for caribou that fall outside the observed range of variation across the country, characterized here (Figure 2) by 95% predictive intervals from a national analysis (Johnson et al. 2020), are implausible. In other words, if we have seen no examples of self-sustaining caribou populations at a particularly high level of disturbance, it is unreasonable to project or expect this outcome. These premises constrain the plausible demographic-disturbance priors, and clarify that no additional information is needed to predict outcomes for caribou where anthropogenic disturbance is high (Figure 5) and no meaningful actions are taken to support recovery.

>>>>

To derive a composition survey bias correction term we assume each group of animals in a calf:cow composition survey contains one or more collared adult females ($T$), and may also include: uncollared adult females misidentified as young bulls or unknown sex ($U$); correctly identified uncollared adult females ($V$); young bulls correctly identified as male or unknown sex ($O$); young bulls misidentified as uncollared adult females ($P$); observed calves ($J$); and unobserved calves ($K$). The apparent number of adult females in the group is $T+V+P=Tw$, where $w$ is the observed ratio of the apparent number of uncollared adult females to collared animals in the recruitment survey. The ratio of young bulls to uncollared adult females in the group is
$q = (P+O)/(U+V).$ Assuming an equal probability $u$ of misidentifying young bulls as adult females and vice versa, we get $V=(U+V)(1-u)$ and $P=(O+P)u$. Given a probability $z$ of missing calves, we get $J=(J+K)(1-z)$. Given these assumptions and definitions, the sex and bias-corrected recruitment rate $X$ can be written as function of the observed calf:cow ratio $R$, the cow multiplier $w$, the ratio of young bulls to adult females $q$, and the misidentification probabilities $u$ and $z$ (see Supplement B for details). For simplicity, we combine the probabilities into a bias correction term $c$ that is integrated into the Bayesian population model and simulations (below): 
$$c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}; X=cR/2$$

@@

We focus on exploring a set of scenarios in which
buffered anthropogenic disturbance, measured as the percentage of area within 500
m of some type of anthropogenic disturbance [@johnson_science_2020; @eccc_scientific_2011],
increases by 1% per year (Figure \@ref(fig:figDisturbance)). The duration of monitoring prior to the projection period varies among scenarios from 1 to 24 years. The projection period begins in 2024, and we consider how uncertainty changes as the projection period increases from 5 to 20 years. The amount of anthropogenic disturbance at the beginning of the projection period varies from low (0\%) through medium-low (20\%) and medium-high (40\%) to high (60\%). These simple hypothetical scenarios span a large range to show how the need for local monitoring to reduce uncertainty depends on landscape state, not to represent the reality of landscape change. Scenarios in which anthropogenic disturbance remains the same or decreases over time are included in Supplement D.

@@

@@

To investigate the effectiveness of various monitoring strategies for reducing uncertainty about population viability, and to show how this varies with landscape condition, we begin with a model for simulating example population dynamics and observations of those dynamics. These example population trajectories are informed by prior information about the distribution of possible parameter values and outcomes from the national analysis of demographic-disturbance relationships [@johnson_science_2020] so we call it the "national" model. Our method for selecting parameter values for each example trajectory yields a set of alternative models (hypotheses) about the state of a particular caribou population; these alternative scenarios informed by prior beliefs are used to assess the expected value of alternative sampling strategies [@runge_which_2011; @dunham_strategic_2023]. We also use the simulated trajectories to confirm that our Bayesian model successfully integrates local data with priors from the national model. For reference, all variables and parameters are defined in supplemental Table A1.

@@

To simulate demographic trajectories for example populations
we used a modified version of Johnson et al's [-@johnson_science_2020]
demographic model described by Dyson et al.
[-@dyson_existing_2022]. Given a population of post-juvenile females at the beginning of year $t$, $\dot{N}_t$, the number that survive to the census, $\dot{W}_t$, is binomially
distributed with true survival probability $\dot{S}_t$:
$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. The calf:cow ratio $\dot{R}_t$ is adjusted for sex ratio, misidentification
biases (Supplement B) and (optionally) a delayed reproduction term to account for the reality that yearlings generally do not breed
[@eacker_web-based_2019; @decesare_estimating_2012] to get maximum potential recruitment of female calves per cow $\dot{X}_t$:
\begin{equation}
  \dot{X}_t=\frac{\dot{c}\dot{R}_t/2}{1+\dot{c}\dot{R}_t/2};\dot{c}=\frac{w(1+\dot{q}\dot{u}-\dot{u})}{(w+\dot{q}\dot{u}-\dot{u})(1-\dot{z})},
  (\#eq:dotX)
\end{equation}
where $w$ is the ratio of observed adult females to collared animals, $\dot{q}$ is the ratio of young bulls to adult females, $\dot{u}$
is the probability of misidentifying young bulls as adult females and
vice versa, and $\dot{z}$ is the probability of missing a calf. We examine sensitivity to these unknown parameters by selecting values for each simulated example population from plausible ranges: $\dot{u}$ and $\dot{v}$ vary uniformly between 0 and 0.2, and $\dot{q}$ varies uniformly between 0 and 0.6.

Realized recruitment rate varies with population density
[@lacy_vortex_2017], and the number of juveniles recruiting to the
post-juvenile class at the census $(\dot{J}_{t})$ is a binomially distributed function
of the number of surviving post-juvenile females and the adjusted
recruitment rate:
\begin{equation}
  \dot{J}_{t} \sim \text{Binomial}(\dot{W}_t,\dot{X}_t[p_0-(p_0-p_k)(\frac{\dot{W}_t}{N_0k})^b]\frac{\dot{W}_t}{\dot{W}_t+a}).
  (\#eq:dotJ)
\end{equation}

@@

As suggested by Dunham et al. [-@dunham_strategic_2023] we also calculate the expected value of perfect information (EVPI) and the expected value of sample information (EVSI) [@runge_which_2011;@yokota_value_2004]. We assign a value or utility of $\nu_{\alpha,\psi}=1$ when the true state of the population $\psi$ matches the assigned a status $\alpha$, and value of $\nu_{\alpha,\psi}=0$ when the status assessment is incorrect. This is "the value ascribed by the decision-maker to the outcome, and is thus the measure of management performance" [@runge_which_2011]. The expected value of current information is the average utility of decisions informed by knowledge of the disturbance scenario $\tau$ (Fig \@ref(fig:figDisturbance)) and priors from the national model:
\begin{equation}
EV_{current} = max_{\alpha}\sum_\psi[P(\psi|\tau)\nu_{\alpha,\psi}],
  (\#eq:EVcurrent)
\end{equation}
where $P(\psi|\tau)$ is the prior probability that the true state of the population is $\psi$, and each population is assigned a status $\alpha$ informed by the prior. The expected value under certainty (i.e. the expected value of a decision when the true state of the population is known) is $EV_{certainty}=1$. The difference between $EV_{certainty}$ and $EV_{current}$ gives EVPI, a measure of the maximum possible benefit that could theoretically be gained if the system were perfectly known.

EVSI measures expected improvement from a particular sampling strategy as the difference between the expected value with and without additional sample information: 
\begin{equation}
EVSI_{\tau} = E_x[max_{\alpha}\sum_\psi[P(\psi|x,\tau)\nu_{\alpha,\psi}]] - EV_{current},
  (\#eq:EVSI)
\end{equation}
where $x$ is the sample information from a particular example trajectory. In this study, we treat simulated observations from 500 example trajectories for each disturbance and monitoring scenario $\tau$ (Figure \@ref(fig:figDisturbance)) as sample from the distribution of possible values of $x$, and assume each of these sample results is equally probable. $P(\psi|x,\tau)$ is the posterior probability that the true state of the population is $\psi$, and each population is assigned a status $\alpha$ informed by the sample information.

@@

Analysis of simulated example demographic trajectories confirms that our Bayesian model can effectively combine local demographic data with priors from national demographic-disturbance relationships to reduce uncertainty in boreal caribou demographic projections (Figs \@ref(fig:examplePredictions) and \@ref(fig:figDiffs)). As expected, uncertainty about the population growth rate decreases as sampling effort in each year (i.e. number of collars) or the number of years of monitoring (Fig \@ref(fig:figDiffs)) increase. Uncertainty is higher when the forecast horizon is long (compare 5 and 20 year projections in Fig \@ref(fig:figDiffs)). Uncertainty is also higher in our high anthropogenic disturbance scenarios (Fig \@ref(fig:figDiffs)) because stochastic variation in population growth rates increase as populations become very small (as in top right panel of Fig \@ref(fig:examplePredictions)).

Our analysis becomes more informative when we examine the probability of correctly assessing population status (Fig \@ref(fig:figStatusErrors)) and the expected value of information (Figs \@ref(fig:figEVSI) and E.1). When anthropogenic disturbance is high (right panels of Figs \@ref(fig:figStatusErrors), \@ref(fig:figEVSI), and E.1)), projections of population status informed only by priors from the national demographic-disturbance relationships are accurate, and adding additional local monitoring data does not improve accuracy. Interestingly, when the projection period is long and anthropogenic disturbance is high, we see a negative EVSI (lower right panel of Fig \@ref(fig:figEVSI)), and we do not see an associated decrease in the probability of correctly assessing population status (lower right panel of Fig \@ref(fig:figStatusErrors)). This means that the long-term projections informed by local monitoring data are on average less precise but not less accurate, which can happen when there is very small prior probability that population is viable (orange band in the top right panel of Fig \@ref(fig:examplePredictions) falls below 1). At lower levels of disturbance, local monitoring can reduce uncertainty, but there are diminishing returns as the number of collars deployed or the number of years of monitoring increases (Figs \@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)). The shapes of the relationships between monitoring effort and the probability of correct status assessment depend on how anthropogenic disturbance is changing over time, but there are diminishing benefits of increased monitoring regardless of whether disturbance is increasing or decreasing (supplement D). Adding unknown bias increases uncertainty, but these increases are not large for the set of parameters we investigate (supplement F).

```{r examplePredictions, fig.cap="Prior (a) and posterior (b) predictions from the Bayesian model for an example population trajectory. In this example scenario anthropogenic disturbance increases by 2\\% per year from 0 to 100\\%. The population is monitored for 16 years with 30 collars per year, and 6 cows per collared cow in the composition surveys. The prior means and 95\\% predictive intervals (top row, orange lines and shading) are similar to the means and ranges between the 2.5\\% and 97.5\\% quantiles of 3000 simulated survival and recruitment trajectories from the national model (top row, blue lines and shading). Local population data (open triangles) reduces uncertainty (bottom row), but when disturbance is very high local data is not required to predict viability (population growth <= 0.99). When projected populations become very small, stochastic variation in projected population growth rate increases (as in the top right panel).", cache=TRUE}

@@

@@

Our model assumes that increasing cumulative anthropogenic disturbance causes a log-linear decrease in caribou demographic rates (equations 4, 5, 10 and 13), reflecting a scientific consensus informed by a number of lines of evidence that cumulative anthropogenic disturbance has a negative impact on boreal caribou [@eccc_scientific_2011; @johnson_science_2020; @gouvernement_du_quebec_revue_2021;@rudolph_demographic_2017;@sorensen_determining_2008;@hervieux_widespread_2013]. Local data can be integrated to estimate local demographic-disturbance parameters that differ from the national means, and in the absence of local data, appropriate priors lead to a range of outcomes consistent with the observed range of variation across the country [Figures 3 and 5 in @johnson_science_2020]. Others [@moeller_integrating_2021; @mcnay_demographic_2022; @eacker_web-based_2019] have also used Bayesian models to estimate caribou demographic rates from local monitoring data, but our addition of prior information from national demographic-disturbance relationships is novel, and important for reducing the risk that lack of local data will be used as a reason to delay conservation action for highly disturbed and poorly monitored populations.  Integrating demographic-disturbance relationships into a viability projection model clarifies that no additional information is needed to project outcomes where disturbance is high and actions to support recovery are not taken.

@@

Variation in anthropogenic disturbance across the Canadian boreal region is partially correlated with variation in climate and primary productivity [@neilson_direct_2022] because forestry occurs in higher productivity areas, and petroleum extraction activities are concentrated in Western Canada. Effects of disturbance on vegetation and foodweb dynamics vary with primary productivity [@neufeld_primary_2021], and some have expressed concern about confounding of demographic-disturbance relationships [@wilson_causal_2021]. One option for addressing these concerns would be to develop regional demographic-disturbance relationships for populations that are ecologically similar to one another (e.g. caribou inhabiting forests that are sufficiently productive for industrial forestry in Eastern Canada) [@eccc_scientific_2024], and use the priors from regional demographic-disturbance relationships to inform demographic projections within the region. However, while acknowledging and allowing for regional variation, we also need to guard against risks that uncertainty will be inflated and used as an excuse to delay action, a tactic that has been used in this and other decision contexts [@bissonnette_controverses_2020; @boan_climate_2018]. In order to be informative and credible, regional demographic-disturbance analyses must include a sufficient number of observations spanning a range from high to low anthropogenic disturbance. Priors that lead to projections of outcomes for caribou that fall outside the observed range of variation across the country, characterized here (Figure 2) by 95% predictive intervals from a national analysis (Johnson et al. 2020), are implausible. In other words, if we have seen no examples of self-sustaining caribou populations at a particularly high level of disturbance, it is unreasonable to project or expect this outcome. These premises constrain the plausible demographic-disturbance priors, and clarify that no additional information is needed to predict outcomes for caribou where anthropogenic disturbance is high (Figure 5) and no meaningful actions are taken to support recovery.

### 2024-06-17 analysis/paper/supplement.Rmd
Updated results. Draft to MH.

knitr::include_graphics(here::here("figs/RecBoth.png"))
```

@@

>>>>

knitr::include_graphics(here::here("figs/RecBoth.png"))
```
\newpage

@@

# Expected value of perfect information

```{r figEVPI, fig.cap="The expected value of perfect information (EVPI) varies among anthropogenic disturbance scenarios (see increasing, decreasing and stable scenarios in Figs 1, D.1 and D.5 respectively) and the number of years projected. When disturbance is high EVPI is low, indicating that local sampling will not alter status assessments. Increasing the forecast horizon can either increase or decrease the EVPI as the anthropogenic disturbance trend makes outcomes more or less predictable.", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/EVPIall.png"))
```
